---
title: "Mini Project 03 - Visualizing and Maintaining the Green Canopy of NYC"
author: "Daniel Ohebshalom"
date: "11-14-2025"
format: 
  html:
    code-fold: true
    theme: pulse
    toc: true
    message: false
    warning: false
    cache: true
---
# Introduction

In America's densest city in terms of population and overall habitation, trees serve a multifaceted purpose in terms of providing clean oxygen, shade and overall comfort to the concrete jungle. This MiniProject explores the distirbution and maintinence of trees across NYC using parks data; visualizing placement patterns and analyzing differences across the city, eventually culminating in a one council proposal.

# Key Packages
```{r}
suppressPackageStartupMessages({
library(sf) 
library(dplyr) 
library(ggplot2)   
library(httr2)        
library(tidyr)
library(plotly)})       
```



# Task 1: Data Acquisition, Downloading NYC Council Boundaries:
```{r}
#| label: download-districts
#| eval: true

download_districts <- function() {
  # Create directory if needed
  dir_path <- "STA9750-2025-FALL/data/mp03"
  if (!dir.exists(dir_path)) {
    dir.create(dir_path, recursive = TRUE)
  }
  
  # Define file paths
  zip_file <- file.path(dir_path, "districts.zip")
  
  # Download if not already present
  if (!file.exists(zip_file)) {
    cat("Downloading NYC City Council District boundaries...\n")
    # URL for NYC Planning City Council Districts (Clipped to Shoreline)
    url <- "https://data.cityofnewyork.us/api/geospatial/erm2-nwe9?method=export&format=Shapefile"
    download.file(url, zip_file, mode = "wb")
    cat("Download complete.\n")
  } else {
    cat("District file already exists. Skipping download.\n")
  }
  
  # The unzipped folder is named nyccwi_25c
  unzipped_dir <- file.path(dir_path, "nyccwi_25c")
  
  # Find and read the shapefile
  shp_file <- file.path(unzipped_dir, "nyccwi.shp")
  
  if (!file.exists(shp_file)) {
    stop("Shapefile not found at ", shp_file)
  }
  
  cat("Reading shapefile:", shp_file, "\n")
  
  districts <- sf::st_read(shp_file, quiet = TRUE)
  
  # Step 2: Transform to WGS84
  districts <- sf::st_transform(districts, crs = "WGS84")
  
  # Optional: Simplify geometry for faster plotting
  districts <- districts |>
    dplyr::mutate(geometry = sf::st_simplify(geometry, dTolerance = 5))
  
  return(districts)
}

# Call the function
districts <- download_districts()

cat("Districts data loaded. Shape:", nrow(districts), "districts\n")
cat("Class:", class(districts), "\n")
head(districts)
```


# Task 2: Downloading Tree Points:
```{r}

#| label: download-trees
#| cache: true

download_trees <- function() {
  # Load required libraries
  library(httr2)
  library(sf)
  library(dplyr)
  library(readr)
  
  # Create directory if needed
  dir_path <- "STA9750-2025-FALL/data/mp03"
  if (!dir.exists(dir_path)) {
    dir.create(dir_path, recursive = TRUE)
  }
  
  # Base API URL for Forestry Tree Points (CSV format via SODA2)
  base_url <- "https://data.cityofnewyork.us/api/views/hn5i-inap/rows.csv"
  
  # Initialize parameters
  limit <- 50000  # Use larger batches for CSV
  offset <- 0
  request_num <- 0
  
  # Iterate through all pages
  repeat {
    request_num <- request_num + 1
    file_name <- file.path(dir_path, paste0("trees_", sprintf("%04d", request_num), ".csv"))
    
    # STOP if we already have 11 files (or your desired amount)
    if (request_num > 11) {
      cat("Reached desired number of files (11). Stopping.\n")
      break
    }
    
    # Check if file already exists
    if (file.exists(file_name)) {
      cat("File already exists:", file_name, "\n")
      
      # Check if it's valid (not blank)
      tryCatch({
        data <- readr::read_csv(file_name, show_col_types = FALSE)
        if (nrow(data) < limit) {
          cat("Last batch detected (", nrow(data), "rows). Stopping.\n")
          break
        }
        offset <- offset + limit
      }, error = function(e) {
        cat("Error reading existing file. File may be corrupted.\n")
      })
      
    } else {
      # File doesn't exist, download it
      cat("Downloading batch", request_num, "with offset", offset, "\n")
      
      # Construct query parameters
      full_url <- paste0(
        base_url,
        "?accessType=DOWNLOAD",
        "&bom=true",
        "&api=true",
        "&limit=", limit,
        "&offset=", offset
      )
      
      # Try to download with error handling
      tryCatch({
        req <- httr2::request(full_url)
        resp <- httr2::req_perform(req, verbosity = 0)
        
        # Check if response is successful
        if (httr2::resp_status(resp) == 200) {
          # Save to file
          writeBin(httr2::resp_body_raw(resp), file_name)
          
          # Verify the file was written and is not blank
          if (file.size(file_name) > 1000) {
            cat("Successfully saved. Size:", 
                format(file.size(file_name), big.mark = ","), "bytes\n")
            
            # Read file to check how many records
            data <- readr::read_csv(file_name, show_col_types = FALSE)
            cat("Records in this batch:", nrow(data), "\n")
            
            # If we got fewer records than the limit, we've reached the end
            if (nrow(data) < limit) {
              cat("Reached end of dataset. Total batches:", request_num, "\n")
              break
            }
            
            offset <- offset + limit
          } else {
            cat("WARNING: File is suspiciously small. Removing...\n")
            file.remove(file_name)
            next
          }
        } else {
          cat("API returned status:", httr2::resp_status(resp), "\n")
          break
        }
        
      }, error = function(e) {
        cat("ERROR downloading batch", request_num, ":", conditionMessage(e), "\n")
        cat("Stopping download process.\n")
      })
      
      # Be respectful: wait between requests
      Sys.sleep(2)
    }
  }
  
  cat("\n=== Download Complete ===\n")
  
  # Read all downloaded CSV files
  cat("Reading all downloaded files...\n")
  all_files <- list.files(dir_path, pattern = "^trees_[0-9]{4}\\.csv$", full.names = TRUE)
  all_files <- sort(all_files)
  
  cat("Found", length(all_files), "files to read.\n")
  
  if (length(all_files) == 0) {
    stop("No tree files found! Download may have failed.")
  }
  
  # Read and combine all files
  trees_list <- list()
  for (i in seq_along(all_files)) {
    cat("Reading file", i, "of", length(all_files), "\n")
    tryCatch({
      trees_list[[i]] <- readr::read_csv(all_files[i], show_col_types = FALSE)
    }, error = function(e) {
      cat("Error reading", all_files[i], ":", conditionMessage(e), "\n")
    })
  }
  
  # Combine all data frames
  cat("Combining all tree data...\n")
  trees <- dplyr::bind_rows(trees_list)
  
  cat("Total trees loaded:", nrow(trees), "\n")
  cat("Columns in data:", paste(names(trees), collapse = ", "), "\n")
  
  # Convert to sf object since it has lat/lon columns
  cat("Converting to spatial data (sf object)...\n")
  
  # Check which columns contain coordinates
  if ("Latitude" %in% names(trees) && "Longitude" %in% names(trees)) {
    trees_sf <- sf::st_as_sf(trees, coords = c("Longitude", "Latitude"), crs = "WGS84")
  } else if ("latitude" %in% names(trees) && "longitude" %in% names(trees)) {
    trees_sf <- sf::st_as_sf(trees, coords = c("longitude", "latitude"), crs = "WGS84")
  } else if ("Location" %in% names(trees)) {
    # Location column contains the geometry - parse it
    cat("Using Location column for geometry...\n")
    trees_sf <- sf::st_as_sf(trees, wkt = "Location", crs = "WGS84")
  } else if ("Geometry" %in% names(trees)) {
    # Geometry column contains the geometry - parse it
    cat("Using Geometry column for geometry...\n")
    trees_sf <- sf::st_as_sf(trees, wkt = "Geometry", crs = "WGS84")
  } else {
    cat("Available columns:", paste(names(trees), collapse = ", "), "\n")
    stop("Could not find coordinate columns in data")
  }
  
  cat("Geometry column created successfully\n")
  
  # Limit to subset for faster processing
  cat("Subsetting to 200,000 trees for faster processing...\n")
  trees_sf <- trees_sf |>
    dplyr::slice_sample(n = 200000)
  
  cat("Final dataset size:", nrow(trees_sf), "trees\n")
  
  return(trees_sf)
}

# Call the function
trees <- download_trees()

cat("Trees data loaded. Total trees:", nrow(trees), "\n")

```


# Task 3: Creating Map Visualization of all Trees across all bouroughs and neighborhoods:

```{r}

#| label: map-all-trees
#| fig-cap: "Distribution of all trees in NYC by council district"

# Create clean ggplot with boundary lines
trees_map <- ggplot2::ggplot() +
  # District boundaries
  ggplot2::geom_sf(
    data = districts,
    fill = "white",
    color = "black",
    linewidth = 0.5
  ) +
  # Tree points
  ggplot2::geom_sf(
    data = trees,
    color = "darkgreen",
    size = 0.8,
    alpha = 0.3,
    shape = 16
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, margin = margin(b = 10)),
    panel.grid.major = element_line(color = "gray90"),
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10)
  ) +
  ggplot2::labs(
    title = "NYC Trees by Council District",
    subtitle = "Distribution of street trees across NYC",
    x = "Longitude",
    y = "Latitude"
  )

print(trees_map)

```

## Spaitially joining trees to districts:

```{r}

#| label: spatial-join

if (!inherits(trees, "sf")) {
  cat("WARNING: trees is not an sf object. Attempting to restore geometry...\n")
  # Recreate sf object from Location column
  trees <- sf::st_as_sf(trees, wkt = "Location", crs = "WGS84")
}

# Spatially join trees to districts
trees_in_districts <- sf::st_join(trees, districts, join = sf::st_intersects)

cat("Trees joined to districts. Total rows:", nrow(trees_in_districts), "\n")
```


# Task 4: District Level analysis:

## Question 1: Which council district has the most trees?



```{r}

#| label: most-trees

trees_per_district <- trees_in_districts |>
  sf::st_drop_geometry() |>
  dplyr::group_by(council_district = CounDist) |>
  dplyr::summarise(n_trees = dplyr::n(), .groups = "drop") |>
  dplyr::arrange(dplyr::desc(n_trees))

most_trees <- trees_per_district |> dplyr::slice(1)

cat("District with most trees: District", most_trees$council_district, 
    "with", most_trees$n_trees, "trees\n")


```


## Question 1: Which council district has the most trees?


```{r}

#| label: most-trees

trees_per_district <- trees_in_districts |>
  sf::st_drop_geometry() |>
  dplyr::group_by(council_district = CounDist) |>
  dplyr::summarise(n_trees = dplyr::n(), .groups = "drop") |>
  dplyr::arrange(dplyr::desc(n_trees))

most_trees <- trees_per_district |> dplyr::slice(1)

cat("District with most trees: District", most_trees$council_district, 
    "with", most_trees$n_trees, "trees\n")


```

## Question 2: Which council district has the highest density of trees? 

```{r}

#| label: tree-density

# Get district areas
district_area <- districts |>
  sf::st_drop_geometry() |>
  dplyr::select(CounDist, Shape_Area)

# Calculate density
trees_density <- trees_in_districts |>
  sf::st_drop_geometry() |>
  dplyr::group_by(council_district = CounDist) |>
  dplyr::summarise(n_trees = dplyr::n(), .groups = "drop") |>
  dplyr::left_join(district_area, by = c("council_district" = "CounDist")) |>
  dplyr::mutate(tree_density = n_trees / Shape_Area) |>
  dplyr::arrange(dplyr::desc(tree_density))

highest_density <- trees_density |> dplyr::slice(1)

cat("District with highest tree density: District", highest_density$council_district, 
    "with", round(highest_density$tree_density, 6), "trees per square unit\n")




```


## Question 3: Which district has highest fraction of dead trees out of all trees?

```{r}

#| label: dead-trees

dead_trees <- trees_in_districts |>
  sf::st_drop_geometry() |>
  dplyr::group_by(council_district = CounDist) |>
  dplyr::summarise(
    n_dead = sum(TPCondition == "Dead", na.rm = TRUE),
    n_total = dplyr::n(),
    frac_dead = n_dead / n_total,
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(frac_dead))

highest_dead_frac <- dead_trees |> dplyr::slice(1)

cat("District with highest dead tree fraction: District", highest_dead_frac$council_district, 
    "with", round(highest_dead_frac$frac_dead * 100, 2), "% dead trees\n")




```

## Question 4: What is the most common tree species in Manhattan?

```{r}

#| label: manhattan-species

trees_with_borough <- trees_in_districts |>
  dplyr::mutate(
    borough = dplyr::case_when(
      CounDist >= 1 & CounDist <= 10 ~ "Manhattan",
      CounDist >= 11 & CounDist <= 18 ~ "Bronx",
      CounDist >= 19 & CounDist <= 32 ~ "Queens",
      CounDist >= 33 & CounDist <= 48 ~ "Brooklyn",
      CounDist >= 49 & CounDist <= 51 ~ "Staten Island",
      TRUE ~ NA_character_
    )
  )

most_common_manhattan <- trees_with_borough |>
  sf::st_drop_geometry() |>
  dplyr::filter(borough == "Manhattan") |>
  dplyr::group_by(GenusSpecies) |>
  dplyr::summarise(n = dplyr::n(), .groups = "drop") |>
  dplyr::arrange(dplyr::desc(n)) |>
  dplyr::slice(1)

cat("Most common tree species in Manhattan:", most_common_manhattan$GenusSpecies, 
    "with", most_common_manhattan$n, "trees\n")




```


## Question 5: What is the species of the tree closest to Baruch’s campus?

```{r}

#| label: baruch-closest-tree

# Create point using the provided function
new_st_point <- function(lat, lon, ...){
  sf::st_sfc(point = sf::st_point(c(lon, lat))) |>
    sf::st_set_crs("WGS84")
}

# Baruch College coordinates: latitude 40.7346, longitude -73.9869
baruch_point <- new_st_point(lat = 40.7346, lon = -73.9869)

# Calculate distance from each tree to Baruch
distance_vector <- sf::st_distance(trees, baruch_point)

trees_distance <- trees |>
  dplyr::mutate(distance = as.numeric(distance_vector)) |>
  dplyr::arrange(distance)

closest_tree <- trees_distance |> dplyr::slice(1)

cat("Species of tree closest to Baruch:", 
    sf::st_drop_geometry(closest_tree)$GenusSpecies, "\n")
cat("Distance:", round(closest_tree$distance, 4), "meters\n")

```


# Task 5 Government Project Design:

District 19 (Flushing, Queens) faces a critical threat from invasive tree pests, particularly the Spotted Lanternfly (SLF) and Asian Longhorn Beetle (ALB), which devastate native tree populations and compromise urban forest health. This proposal requests dedicated Parks Department funding to implement a comprehensive invasive species management program focused on early detection, treatment, and removal of infested trees. The program will include targeted pesticide application, removal of high-risk trees, and replanting with pest-resistant native species.

Proposed Scope: ~500 trees requiring invasive species treatment/removal over 18 months

## District Selection (19):

```{r}

#| label: my-district-analysis

my_district <- 19

# Filter data for district 19
my_district_trees <- trees_in_districts |>
  dplyr::filter(`CounDist` == my_district)

my_district_boundary <- districts |>
  dplyr::filter(`CounDist` == my_district)

cat("Your district has", nrow(my_district_trees), "trees\n")

```

## Map of Selected District:

```{r}

#| label: district-zoomed-map
#| fig-cap: "Trees in your selected district"

district_map <- ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = my_district_boundary,
    fill = "lightgray",
    color = "black"
  ) +
  ggplot2::geom_sf(
    data = my_district_trees,
    color = "darkgreen",
    size = 1.5,
    alpha = 0.6
  ) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = paste("Trees in Council District", my_district),
    x = "Longitude",
    y = "Latitude"
  )

print(district_map)

```


## District 19 Quantitiative Comparisions:

```{r}

#| label: district-19-comparison

# Select District 19 and comparison districts (18, 20, 21, 22)
comparison_dists <- c(19, 18, 20, 21, 22)

# Get tree counts and density for comparison
district_comparison <- trees_in_districts |>
  sf::st_drop_geometry() |>
  dplyr::filter(CounDist %in% comparison_dists) |>
  dplyr::group_by(CounDist) |>
  dplyr::summarise(
    tree_count = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    districts |> sf::st_drop_geometry() |> dplyr::select(CounDist, Shape_Area),
    by = "CounDist"
  ) |>
  dplyr::mutate(
    tree_density = tree_count / Shape_Area,
    district_name = dplyr::case_when(
      CounDist == 19 ~ "District 19 (Flushing)",
      CounDist == 18 ~ "District 18 (Astoria)",
      CounDist == 20 ~ "District 20 (Bayside)",
      CounDist == 21 ~ "District 21 (Fresh Meadows)",
      CounDist == 22 ~ "District 22 (Whitestone)",
      TRUE ~ paste("District", CounDist)
    )
  ) |>
  dplyr::arrange(dplyr::desc(tree_density))

print(district_comparison)

cat("\n=== District 19 Analysis ===\n")
d19_data <- district_comparison |> dplyr::filter(CounDist == 19)
cat("Total trees in District 19:", d19_data$tree_count, "\n")
cat("Tree density:", round(d19_data$tree_density, 6), "trees per square unit\n")
cat("Rank by density: #", which(district_comparison$CounDist == 19), "out of", nrow(district_comparison), "\n")

```


## Tree Density Analysis: Comparative Visualization:


```{r}

#| label: density-comparison-plot
#| fig-cap: "Tree Density Comparison: District 19 vs. Neighboring Districts"

density_plot <- district_comparison |>
  dplyr::arrange(tree_density) |>
  ggplot2::ggplot(aes(x = reorder(district_name, tree_density), y = tree_density, fill = CounDist == 19)) +
  ggplot2::geom_col() +
  ggplot2::scale_fill_manual(
    values = c("FALSE" = "lightblue", "TRUE" = "darkred"),
    guide = "none"
  ) +
  ggplot2::coord_flip() +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = element_text(size = 13, face = "bold"),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10)
  ) +
  ggplot2::labs(
    title = "Tree Density by District",
    subtitle = "District 19 highlighted in red",
    x = "District",
    y = "Tree Density (trees per square unit)"
  )

print(density_plot)

```

District 19 has significant tree canopy across its boundaries. However, its relatively high tree density makes it particularly vulnerable to invasive pest outbreaks. Unlike districts with lower densities, District 19's concentrated tree population in a mostly suburban landscape creates ideal conditions for rapid pest spread. This density makes it an urgent priority for preventive invasive species management.


## District Comparison: Tree Health Considerations

```{r}
#| label: tree-health-comparison

# Analyze tree condition across districts
health_comparison <- trees_in_districts |>
  sf::st_drop_geometry() |>
  dplyr::filter(CounDist %in% comparison_dists) |>
  dplyr::group_by(CounDist) |>
  dplyr::summarise(
    total_trees = dplyr::n(),
    dead_trees = sum(TPCondition == "Dead", na.rm = TRUE),
    poor_condition = sum(TPCondition == "Poor", na.rm = TRUE),
    pct_dead = (dead_trees / total_trees) * 100,
    pct_poor = (poor_condition / total_trees) * 100,
    .groups = "drop"
  ) |>
  dplyr::mutate(
    district_name = dplyr::case_when(
      CounDist == 19 ~ "(Flushing)",
      CounDist == 18 ~ "(Astoria)",
      CounDist == 20 ~ "(Bayside)",
      CounDist == 21 ~ "(Fresh Meadows)",
      CounDist == 22 ~ "(Whitestone)",
      TRUE ~ paste("District", CounDist)
    )
  )

print(health_comparison)


```

## Tree Condition Comparison Visualization

```{r}

#| label: condition-comparison-plot
#| fig-cap: "Percentage of Dead/Poor Condition Trees: District 19 vs. Neighboring Districts"

condition_long <- health_comparison |>
  dplyr::select(district_name, pct_dead, pct_poor) |>
  tidyr::pivot_longer(
    cols = c(pct_dead, pct_poor),
    names_to = "condition_type",
    values_to = "percentage"
  ) |>
  dplyr::mutate(
    condition_type = dplyr::recode(
      condition_type,
      pct_dead = "Dead",
      pct_poor = "Poor Health"
    ),
    is_d19 = grepl("District 19", district_name)
  )

condition_plot <- condition_long |>
  ggplot2::ggplot(aes(x = district_name, y = percentage, fill = condition_type)) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::scale_fill_manual(
    values = c("Dead" = "darkred", "Poor Health" = "orange"),
    name = "Tree Condition"
  ) +
  ggplot2::coord_flip() +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = element_text(size = 13, face = "bold"),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10)
  ) +
  ggplot2::labs(
    title = "Tree Health Status: Dead and Poor Condition Trees",
    x = "District",
    y = "Percentage of Trees (%)"
  )

print(condition_plot)

```

District 19's  percentage of dead and poor-condition trees suggests ongoing pest pressure, mainly due to a high suburban sprawl and numerous greenspaces. This proposal targets removal of  compromised trees in affected areas while implementing preventive measures on healthy specimens before pest infestation becomes irreversible.

## Implementation Map of Proposal:  District 19 vs District 20

```{r}

#| label: district-comparison-map
#| fig-cap: "Side-by-Side Comparison: District 19 (proposed) vs. District 20 (reference)"

# Verify trees is sf object, convert if needed
if (!inherits(trees, "sf")) {
  cat("Converting trees to sf object...\n")
  trees <- sf::st_as_sf(trees)
}

# Spatially join trees to get district info for filtering
trees_with_district <- sf::st_join(trees, districts, join = sf::st_intersects)

# Get District 20 trees
district_20_trees <- trees_with_district |>
  dplyr::filter(CounDist == 20)

district_20_boundary <- districts |>
  dplyr::filter(CounDist == 20)

# Create comparison map
comparison_map <- ggplot2::ggplot() +
  # District 19
  ggplot2::geom_sf(
    data = my_district_boundary,
    fill = "white",
    color = "darkred",
    linewidth = 1
  ) +
  ggplot2::geom_sf(
    data = my_district_trees,
    color = "forestgreen",
    size = 1.2,
    alpha = 0.4
  ) +
  # District 20
  ggplot2::geom_sf(
    data = district_20_boundary,
    fill = "white",
    color = "navy",
    linewidth = 1
  ) +
  ggplot2::geom_sf(
    data = district_20_trees,
    color = "blue",
    size = 1.2,
    alpha = 0.4
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "gray90")
  ) +
  ggplot2::labs(
    title = "District 19 (red) vs. District 20 (blue): Tree Distribution",
    subtitle = "District 19 shows comparable tree density with higher vulnerability to invasive species.",
    x = "Longitude",
    y = "Latitude"
  )

print(comparison_map)

```

Conclusion:

District 19's combination of high tree density, elevated levels of dead/poor-condition trees, and geographic proximity to major invasive species makes it an ideal candidate for this specialized program. The proposal will fund removal of 200 compromised trees, preventive treatment of 150 vulnerable specimens, and monitoring of 150 at-risk trees over 18 months in the area. This proactive investment will protect District 19's urban forest and preserve the green canopy this community depends on for their livelihood.
 